Discussion

### 2020-03-14 doc/Discussion.rmd
notes for outline draft

@@

@@

The Sardel guppies paper was able to deduce (dientangle) that this typical landscape pattern was due to // disentangle the specific components of centromere supression and telomere bias between the male and feamel landscapes. we are not able to do that in these cases (because of inability to estimate the chromosome identity across our sample of single bivalents. )

@@

@@

  it boils down to, the relationship with axis length and DNA loop size -- 
  --females have shorter loops - longer axis and weaker interference (in abstracted sense)

@@

@@

 mouse and humans show greater between cell variance 
 this metric isn't really focused on (not many studies that examine inidivial level and both sexes)
3. Within mouse variance differences
In our results we observe more between cell variance in females (across all strains). This pattern has been observed in humans (Lynn et al). (But results of reporting observations of single meiosis in both sexes).

 Our results confirm many of the typical result fir the sex difference in the recombination landscape.

 For many years the canoncial pattern of heterochiasmy has been dicribed (with linkage maps) as females have more genome wide recombination rates compared to males of the sames speces (ref older hetC refs)

more recntly more nuanced patterns have also been noted/discribed in greater detail; general/broad scale landscape patterns male telomeric and female uniform, and chromatin compaction (in some species)

@@

@@

1. More DSBs in the high rec strains
2. The conserved ratios of DSB : CO implies that the increase in DSB numbers are accompanied by proportional increase in crossovers. This is evidence that shifts support away from the CO:NCO decision to an earlier prophase stage such as when the chromatin compaction is set up. 
Together these results indicate that (it's not the CO : NCO decision but an earlier step (when the total number of chromatin loops (the substrate for SPO11, are laid down with the establishment of the meiotic chromosome structure (DNA loop size)

@@

3. DSB number being driven by the number of chromatin loops which is positively correlated with axis length.
4. These results supported by the moderate support of longer SC length (pachytene stage in the high rec males).
(Review Baier et al 2014)

@@

##3.Two-locus (protect for drivers)
Asymmetric division of egg opens them up to meiotic drivers. Recombination modifiers for sex specific female landscape will decrease the chances of driving centromeres to segregate to the egg by increasing the number of crossovers and placing them across a larger area of the genome (uniform placement).

## 4. Sindle   // Difference in acentromeric spindle and SAC
A fundamental difference in meiosis (MI) for males and females is the presence of the centrosome which serves as a nucleation site for MT at each pole (and changes the shape of the spindle). (Our hypothesis is that the tension in a centrosome spindle is stronger (or more uniform) since all MT-KT are anchored at single points across cells (the two centrosomes). This will also make prediction for the spindle assembly checkpoint (SAC), which is stronger/stricter in smaller gametes with centrosomes.  (the SAC is more sensitive to achiasmate bivalents on the spindle / better at detecting a lack of tension).
Whereas in the acentrosome spindle is more diffuse (across a larger area) and the tension across bivalents are anchored by multiple MTOCs, resulting in weaker tension force or less uniform strengths across all the bivalents within the cells. The SAC in eggs seems to be weaker/leakier, that is more achiasmate bivalents are required to trigger the SAC (stop the division).

<schuh ellenberg 2007 spindle

@@

bivalent orientation early metaphase -- search and capture model -- when there are mulitple KT-MT attachmments, (lateral (on the side or not on centromere/KT proper), merotelic MT attachemtn from ooposite poles) and amplitelic (MT from same pole)
then biorientation process starts, --search and capature process (only amplictic Mt-KT attachments get stablize)

>>>>

# Differences across models

Explain the model differences;

## 1.can it be applied to broad scale chromosome level landscape?

-genome wide recombination rates are made of the chromosome level patterns

This is important because general patterns have been observed for the recombination landscape at the chromosome level (also small range of crossovers per chromosome and broad scale pattern (Haenel et al ))

## 2. Does the model account for evolution?

given all the variance observed and fast evolution, models have to have some engine for evolution.

(some source of affecting fitness, or shifting the optima)

@@

These are the metrics for the _conserved sex differences_

@@

The Sardel guppies paper was able to deduce (dientangle) that this typical landscape pattern was due to // disentangle the specific components of centromere supression and telomere bias between the male and feamel landscapes. we are not able to do that in these cases (because of inability to estimate the chromosome identity across our sample of single bivalent
s. )

@@

1. SC length  2. connection to interference

general pattern for the sexual dimorphsim the relationship between axis length and CO number
-- axis are longer in females than male, suggesting suport more crossovers (more DSBs)
but the density is less in females.

Our results from MSM and PWD show that the sc length compaction is independant (of the direction of heterochiasmy. Two strains have have more crossovers in males, but females still have longer sc axis. _This indecates that the sexual dimorphism in chromosome organization / DNA compaction is a conserved part of recombination landscape_

The pattern of interference across raw and normalized measures -- also confirms previous results of male having stronger interfere -- due to the longer loops (Petkov)

@@

  it boils down to, the relationship with axis length and DNA loop size --

  --females have shorter loops - longer axis and weaker interference

@@

  same pattern in humans (Gruhn et al 2013)

@@

- Our results confirm previous findings of females having greater between cell variation (for indiviual females compared to males)
the results -- indicate there is more than can be due to technical noise (for oocytes vs spermatocytes spreads)
also all ovary collections were from early/ young mice -- not due to the maternal age effects, (which has to do with the stability of bivalents at metaphase -- after sexual maturity and ovulation (ref))

We did not observe

- Because these are inbred strains raised in common / standard lab environment, (an single levels)

_sources of within animal variation_

-age differences

new paper has revealed within nucleus variation for (CO number -- potentially via the chromatin remodeling) (Veller Kleckner)

prev reported sex diffferences in

 Gruhn et al 2013 (the favorite figures)

 Gruhn et al 2016 -- not due to the developmental stage of the oocytes (ovulation) (the production line)

 mouse and humans show greater between cell variance -- this metric isn't really focused on (not many studies that examine inidivial level and both sexes)

3. Within mouse variance differences
In our results we observe more between cell variance in females (across all strains). This pattern has been observed in humans (Lynn et al). (But results of reporting observations of single meiosis in both sexes).

(references, more within animal variance in
(number of crossovers per cell)
but maybe, rec landscapes in general)

(Gruhn et al), Kong et al?

cattle,

- age effects?

- Hassold hunt

_Relaxed selection_ leads to

@@

Our results find evolution between a subset of strains -- demonstrating that the high rec strains have greater DSB number

@@

1. More DSBs in the high rec strains

2. The conserved ratios of DSB : CO implies that the increase in DSB numbers are accompanied by proportional increase in crossovers. This is evidence that shifts support away from the CO:NCO decision to an earlier prophase stage such as when the chromatin compaction is set up. 
Together these results indicate that (it's not the CO : NCO decision but an earlier step (when the total number of chromatin loops (the substrate for SPO11, are laid down with the establishment of the meiotic chromosome structure (DNA loop size)

@@

3. DSB number being driven by the number of chromatin loops which is positively correlated with axis length.
4. These results supported by the moderate support of longer SC length (pachytene stage in the high rec males).

_the resuts support / replicate those reported in Baier et al_

a study of in house mouse -- using different markers for DSBs and intermediates (markers for the recombination pathway (from DSB to mature crossover))

-different DSB marks, distinct inbred strains two classical lab strains and a wild derived strain from castenus, used FISH for chromosome specific measures

our results are similar in several ways
- correlation of final MLH1 and DBS counts, SC measues correlate with the DSB number (even tho the differnces we observe in our study are -- slight / less significant -- we attribute that to chromosome specific noise )

the correlation of DSB marker with MLH1

(Review Baier et al 2014)

@@

-- this model is an extension of the modifer model presented in Lenormand

-- (basics)
 extends the lenomand focusing on two conditions. first that the stength of epistatsis in dependant on the cis or trans nature. the authors note that this would apply to gene archtechture (and increase with in cis ) -- where the regulatroy region and coding region of a gene evolve (together)

 second that the relative interference strengths differnt between male and female.  This can apply to (genes acting in sexual antagonism) or can be mediated via imprinted genes.

This new model has _strengths_

  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

 - incorporates know cellular / molecular mechanisms with a very common pattern sexual dimorphic recombination rate landscape.

- extends modifer model to explain a chromosome based / level pattern

_weaknesses_, and ways the model diverges from our results

 - the simulation present in the paper are too weak to support the rapid spread of such a modifer system (selection coefficients are too weak to support the evolution -- spread of such a modifer.)

- Other ways this model departs from our results,

-Always predicting male will be lower recombination rates (or risk breaking up the epistatsis blocks) -- which doesn't fit with our results of faster male specific evolution in gwRR.

However / while not explicitaly arameterized or predicted in this model,
 -- this model predicts larger chuncks of chromosomes will segregate together
 (as outlined in Veller et al)

 our results of stronger male interference and stronger interference in high rec strains have the same consequences as predited from this modle

##3.Two-locus (protect for drivers)

Asymmetric division of egg opens them up to meiotic drivers. Recombination modifiers for sex specific female landscape will decrease the chances of driving centromeres to segregate to the egg by increasing the number of crossovers and placing them across a larger area of the genome (uniform placement).

_model basics_

'break up driv system consiracies'

_strengths_

a clear fitness effect (the drive)

explains female landscape pattern in reference to the centromere (presummed location of drivers), yet the model doesn't scale down to chromsome level (such as to explain, chromosome size differences)

_weaknesses_
(modifers - driver systems everywhere!), impossible to verify, doesn't account for male landscape pattern (supression from centromere to end)

## 4. Spindle   // Difference in acentromeric spindle and SAC

A fundamental difference in meiosis (MI) for males and females is the presence of the centrosome which serves as a nucleation site for MT at each pole (and changes the shape of the spindle). (Our hypothesis is that the tension in a centrosome spindle is stronger (or more uniform) since all MT-KT are anchored at single points across cells (the two centrosomes). This will also make prediction for the spindle assembly checkpoint (SAC), which is stronger/stricter in smaller gametes with centrosomes.  (the SAC is more sensitive to achiasmate bivalents on the spindle / better at detecting a lack of tension).
Whereas in the acentrosome spindle is more diffuse (across a larger area) and the tension across bivalents are anchored by multiple MTOCs, resulting in weaker tension force or less uniform strengths across all the bivalents within the cells. The SAC in eggs seems to be weaker/leakier, that is more achiasmate bivalents are required to trigger the SAC (stop the division).

we like this model because it can be applied to _both our results_ for i) sexual dimorphism variation and ii) the fast male evolution in gwRR and recombination landscape.

for the distinct _evolutionary patterns_ selection on the recombination landscape via the spindle (segregation checkpoint),

_main sex differences in spindle_ are the anchoring of MT at opposite poles

in males there are centromsomes for Mi and MII divisions
where are in oocytes, a (universal) conserved feature of oogenesis in many species is that they loose centrosomes -- instead the MTs are anchored at mulitple aMTOCs.

There is potential for these aMTOC to be unevenly distributd across the cortical and central side of the spindles (Wu et al, Schuh and ellenberg) --- (could be an origin of different tension / pulling )

The survalence of SAC in oocytes might be decreased due to large volumne difusing the signals.

_leaky SAC_ for oocytes (is known) .. meaning that bivalents / homologs without proper tension force

(potentially this bad checkpoint is adaptive for thwarting centromere drive systems, which work through the SAC and tension signals)

how the SAC spindle works, correctly

evidence that the oocyte spindle is leaky

_precition that the leaky spindle makes relaxes selection_

aTMOC's  (because there are mulitple, these could be uneven across the cell)

_asymetrical division_

this makes egg weak to driving centromeres which (are segregation distortors), that means some centromeres segreation (rotate to the polar body side of the egg)

_number of placement of crossovers influencing spindle_

For this model we propose that the number and placement of crossovers -- will affect the

_strengths of model_

_weakness_

No very few spermatocyte spindle data  (it's assumed to )

@@

References:
Schuh Ellenberg 2007 spindle

## notes for Wu thesis
bivalent orientation early metaphase -- search and capture model -- when there are mulitple KT-MT attachmments, (lateral (on the side or not on centromere/KT proper), merotelic MT attachemtn from ooposite poles) and amplitelic (MT from same pole)
then biorientation process starts, --search and capature process (only amplictic Mt-KT attachments get stablize)
